# Generating Effect Sizes: `generate_smd()` {#sec-generate-smd}

```{r}
#| label: setup-smd
#| include: false
source("_common.R")
library(mvtnorm)
```

```{r}
#| label: source-functions
#| include: false
source("R/simulation-functions.R")
```

In the previous chapter we defined the working models that describe *between-study* dependence. This chapter turns to a different question: how do we **generate** the effect size estimates that those models will analyze? The function `generate_smd()` is the workhorse that produces standardized mean difference (SMD) effect sizes for a single study. Understanding it requires some background on what SMDs are, how they relate to raw data, and why the simulation can skip the raw-data step entirely.


## What are standardized mean differences? {#sec-smd-background}

A standardized mean difference compares two group means on a common, unit-free scale by dividing the raw mean difference by a pooled standard deviation. Two variants appear throughout the meta-analysis literature.

**Cohen's $d$.** The most direct estimate divides the observed mean difference by the pooled within-group standard deviation:

$$
d = \frac{\bar{Y}_T - \bar{Y}_C}{S_p}
$$ {#eq-cohens-d}

where $S_p$ is the pooled standard deviation computed from both groups.

**Hedges' $g$.** Cohen's $d$ has a slight positive bias in small samples. @hedges1981 introduced a correction factor $J$ that removes this bias:
$$
g = J \times d, \qquad J = 1 - \frac{3}{4(N - 2) - 1}
$$ {#eq-hedges-g}

where $N$ is the total sample size across both groups. As $N$ grows, $J \to 1$ and the correction vanishes. For a study with $N = 20$, $J \approx 0.96$; for $N = 200$, $J \approx 0.996$.

SMDs are the dominant effect size metric in education, psychology, and the behavioral sciences because they allow researchers to combine studies that measured the same construct on different scales. In the @pustejovsky2022 simulation, every generated effect size is a Hedges' $g$.


## The statistical model for generating effect sizes {#sec-stat-model}

Imagine a single study $j$ that compares a treatment group and a control group on $n_j$ different outcomes. The study has total sample size $N_j$, split equally into two groups of $N_j / 2$.

### Multivariate outcomes with equi-correlation

The $n_j$ outcomes within a study are not independent --- a student who scores high on one reading measure probably scores high on another. The simulation captures this within-study dependence through an **equi-correlation** structure. The population covariance matrix of the $n_j$ outcomes is

$$
\boldsymbol{\Sigma}_j = \rho_j \mathbf{J}_{n_j} + (1 - \rho_j)\mathbf{I}_{n_j}
$$ {#eq-sigma-equicorr}

where $\mathbf{J}_{n_j}$ is the $n_j \times n_j$ matrix of ones, $\mathbf{I}_{n_j}$ is the identity matrix, and $\rho_j$ is the common correlation among all outcome pairs in study $j$.

Why equi-correlation? It is a deliberate simplifying assumption. Real within-study correlations vary across outcome pairs, but specifying a full $n_j \times n_j$ correlation matrix for every study would be impractical and would add parameters that the paper is not trying to study. The equi-correlation structure keeps the data-generating process manageable while still inducing realistic dependence among the effect sizes within each study.

::: {.callout-note}
## Standardized covariance matrix
Because the outcomes are standardized (both groups share the same pooled standard deviation on each outcome), $\boldsymbol{\Sigma}_j$ is simultaneously a correlation matrix and a covariance matrix. Every diagonal entry is 1 and every off-diagonal entry is $\rho_j$.
:::


## Step-by-step derivation {#sec-derivation}

The simulation generates Hedges' $g$ and its sampling variance without ever creating individual-level data. Instead, it works with the sufficient statistics that $d$ and $g$ depend on: the vector of mean differences and the pooled covariance matrix. Here is the five-step derivation, following the supplementary materials of @pustejovsky2022.

### Step 1: Unstandardized mean differences

For two groups each of size $N_j / 2$, the vector of unstandardized mean differences $\mathbf{D}_j$ has a multivariate normal distribution:

$$
\mathbf{D}_j \sim N\!\left(\boldsymbol{\delta}_j, \; \frac{4}{N_j}\boldsymbol{\Sigma}_j\right)
$$ {#eq-mean-diff}

The factor $4 / N_j$ comes from the variance of a difference of two means, each based on $N_j / 2$ observations: $\text{Var}(\bar{Y}_T - \bar{Y}_C) = \sigma^2 (1/(N_j/2) + 1/(N_j/2)) = 4\sigma^2 / N_j$, and the $\sigma^2$ is absorbed into $\boldsymbol{\Sigma}_j$ (which has unit diagonal).

### Step 2: Pooled sample covariance matrix

The pooled sample covariance matrix $\mathbf{S}_j$ follows a Wishart distribution:

$$
(N_j - 2) \times \mathbf{S}_j \sim \text{Wishart}(N_j - 2, \; \boldsymbol{\Sigma}_j)
$$ {#eq-wishart}

The degrees of freedom are $N_j - 2$ because pooling two groups of size $N_j/2$ yields $N_j/2 - 1 + N_j/2 - 1 = N_j - 2$ degrees of freedom. The diagonal elements $s^2_{ij}$ of $\mathbf{S}_j$ serve as the denominators for the SMDs.

### Step 3: Cohen's $d$

Each outcome's Cohen's $d$ is the ratio of the mean difference to the square root of the corresponding diagonal element of the pooled covariance:

$$
d_{ij} = \frac{D_{ij}}{\sqrt{s^2_{ij}}}
$$ {#eq-cohens-d-gen}

### Step 4: Hedges' $g$ (bias correction)

Apply the small-sample correction factor:

$$
g_{ij} = J \times d_{ij}, \qquad J = 1 - \frac{3}{4(N_j - 2) - 1}
$$ {#eq-hedges-g-gen}

::: {.callout-tip}
## Equivalent expressions for $J$
The paper writes the correction as $J = 1 - 3/(4N_j - 9)$. Expanding the code's version: $4(N_j - 2) - 1 = 4N_j - 8 - 1 = 4N_j - 9$. They are identical.
:::

```{r}
#| label: fig-correction-factor
#| fig-cap: "The small-sample correction factor J as a function of total sample size N."
#| fig-width: 7
#| fig-height: 4

N_seq <- seq(10, 500, by = 1)
J_seq <- 1 - 3 / (4 * (N_seq - 2) - 1)

ggplot(data.frame(N = N_seq, J = J_seq), aes(x = N, y = J)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  annotate("point", x = 20, y = 1 - 3/(4*18-1), color = "firebrick", size = 3) +
  annotate("text", x = 25, y = 1 - 3/(4*18-1) - 0.003,
           label = sprintf("N = 20: J = %.3f", 1 - 3/(4*18-1)),
           hjust = 0, color = "firebrick", size = 3.5) +
  annotate("point", x = 200, y = 1 - 3/(4*198-1), color = "firebrick", size = 3) +
  annotate("text", x = 210, y = 1 - 3/(4*198-1) + 0.002,
           label = sprintf("N = 200: J = %.4f", 1 - 3/(4*198-1)),
           hjust = 0, color = "firebrick", size = 3.5) +
  scale_y_continuous(limits = c(0.94, 1.001)) +
  labs(x = "Total sample size (N)", y = "Correction factor (J)")
```

The correction is most pronounced for small studies. By $N = 100$, the bias is less than 1%.

### Step 5: Sampling variance of $g$

The approximate sampling variance of each Hedges' $g$ is:

$$
V_{ij} = J^2 \cdot \frac{4}{N_j} + \frac{g_{ij}^2}{2(N_j - 2)}
$$ {#eq-var-g}

::: {.callout-note}
## Realized vs true parameter in $V_{ij}$
The variance formula uses the *realized* effect size $g_{ij}$ (the observed, random value), not the true parameter $\delta_{ij}$. This is important: in the simulation, the sampling variance $V_{ij}$ is computed from the generated data, not from the true effect sizes. Students sometimes erroneously use the true $\delta$ to compute weights for the meta-analysis --- this would create a dependency between the weights and the data-generating parameters that does not exist in practice.
:::

The first term, $J^2 \cdot 4/N_j$, reflects the bias-corrected variance of the mean difference. The second term, $g_{ij}^2 / [2(N_j - 2)]$, reflects the additional uncertainty from estimating the standard deviation in the denominator. Note that $J^2$ multiplies only the first term, and the second term uses $g_{ij}^2$ (Hedges' $g$, not Cohen's $d$). Together, they give the large-sample approximation to the variance of a standardized mean difference.


## The `generate_smd()` function {#sec-function-walkthrough}

Here is the complete function from `R/simulation-functions.R`:

```{r}
#| label: generate-smd-code
#| eval: false
#| code-fold: true
#| code-summary: "Full generate_smd() function"
generate_smd <- function(delta, k, N, Sigma) {

  # make sure delta is a vector
  delta_vec <- rep(delta, length.out = k)

  # create Sigma matrix assuming equicorrelation
  if (!is.matrix(Sigma)) Sigma <- Sigma + diag(1 - Sigma, nrow = k) # cor matrix for 1 study

  # generate numerator of SMD
  mean_diff <- rmvnorm(n = 1, mean = delta_vec, sigma = (4/N) * Sigma)

  # covariance
  cov_mat <- as.matrix(rWishart(n = 1, df = N - 2, Sigma = Sigma)[,,1])
  sigma_sq <- diag(cov_mat) / (N - 2)

  # SMD
  d <- as.vector(mean_diff / sqrt(sigma_sq))  # cohen's d
  J <- (1 - (3/((4 * (N - 2)) - 1)))
  g <- d * J # Hedges g
  var_g <- J^2 * (4 / N + d^2 / (2 * (N - 2)))

  dat <- tibble(g = g, var_g = var_g)

  return(dat)
}
```

Let us walk through the function argument by argument and line by line.

**Arguments:**

| Argument | Description |
|:---------|:------------|
| `delta`  | Vector of true effect sizes $\boldsymbol{\delta}_j$ (length $k$) |
| `k`      | Number of outcomes $n_j$ in this study |
| `N`      | Total sample size $N_j$ |
| `Sigma`  | Either a scalar $\rho_j$ or a full $k \times k$ covariance matrix |

**Line-by-line annotation:**

```r
delta_vec <- rep(delta, length.out = k)
```
Ensures `delta` has exactly `k` elements. If a scalar is passed (e.g., a single true effect size shared across all outcomes), it gets recycled to length `k`.

```r
if (!is.matrix(Sigma)) Sigma <- Sigma + diag(1 - Sigma, nrow = k)
```
If `Sigma` is a scalar (the common correlation $\rho_j$), this constructs the equi-correlation matrix from @eq-sigma-equicorr. The expression `Sigma + diag(1 - Sigma, nrow = k)` creates a matrix with $\rho_j$ on the off-diagonals and 1 on the diagonal.

```r
mean_diff <- rmvnorm(n = 1, mean = delta_vec, sigma = (4/N) * Sigma)
```
Draws the mean difference vector $\mathbf{D}_j$ from @eq-mean-diff. The `rmvnorm()` function from the **mvtnorm** package generates one draw from a multivariate normal with mean $\boldsymbol{\delta}_j$ and covariance $(4/N_j)\boldsymbol{\Sigma}_j$.

```r
cov_mat <- as.matrix(rWishart(n = 1, df = N - 2, Sigma = Sigma)[,,1])
sigma_sq <- diag(cov_mat) / (N - 2)
```
Draws $(N_j - 2) \times \mathbf{S}_j$ from the Wishart distribution (@eq-wishart), then recovers $s^2_{ij}$ by dividing the diagonal by $N_j - 2$. The `[,,1]` indexing extracts the single matrix from the 3D array that `rWishart()` returns.

```r
d <- as.vector(mean_diff / sqrt(sigma_sq))
```
Computes Cohen's $d$ for each outcome (@eq-cohens-d-gen).

```r
J <- (1 - (3/((4 * (N - 2)) - 1)))
g <- d * J
```
Applies the Hedges' $g$ correction (@eq-hedges-g-gen).

```r
var_g <- J^2 * (4 / N + d^2 / (2 * (N - 2)))
```
Computes the approximate sampling variance of each $g_{ij}$ (@eq-var-g). Note that in the code, `J^2` multiplies only `4 / N`, while the second term uses `d^2` rather than `g^2`. This is algebraically equivalent to the formula in @eq-var-g because $g = Jd$, so $g^2 / [2(N-2)] = J^2 d^2 / [2(N-2)]$, and factoring $J^2$ out of both terms recovers the code's expression.

```r
dat <- tibble(g = g, var_g = var_g)
```
Returns a tidy tibble with one row per outcome, containing the effect size estimate and its sampling variance.


## Understanding the Wishart distribution {#sec-wishart}

::: {.callout-note}
## The Wishart distribution: a primer for social scientists

If the chi-squared distribution describes the variability of a single sample variance, the **Wishart distribution** describes the variability of an entire sample covariance matrix. Formally, if you draw $\nu$ independent vectors from a multivariate normal distribution with covariance $\boldsymbol{\Sigma}$, the sum of their outer products follows a $\text{Wishart}(\nu, \boldsymbol{\Sigma})$ distribution.

**If this is new to you**, think of it as follows. Imagine you collected raw data from $N_j / 2$ treatment participants and $N_j / 2$ control participants, each measured on $k$ outcomes. You would compute a $k \times k$ pooled covariance matrix from that data. The Wishart distribution describes exactly how that covariance matrix would vary across hypothetical replications of the study. By drawing directly from the Wishart, we get the same pooled covariance matrix we *would* have gotten from raw data --- without needing to simulate every individual data point.

**Why not simulate raw data?** You certainly could. Generate $N_j/2$ multivariate normal vectors for each group, compute group means and pooled covariance, then form $d$ and $g$. The result would be statistically identical. But the Wishart shortcut is:

- **Faster.** One matrix draw replaces $N_j$ vector draws plus a covariance computation.
- **Cleaner.** The code directly mirrors the statistical model.
- **Standard practice.** This is the approach used in @pustejovsky2022 and in most simulation studies of SMDs.

The key identity is: if $\mathbf{Y}_1, \ldots, \mathbf{Y}_\nu \stackrel{iid}{\sim} N(\mathbf{0}, \boldsymbol{\Sigma})$, then $\sum_{i=1}^{\nu} \mathbf{Y}_i \mathbf{Y}_i' \sim \text{Wishart}(\nu, \boldsymbol{\Sigma})$.
:::

```{r}
#| label: fig-wishart-demo
#| fig-cap: "Distribution of diagonal elements from Wishart-distributed matrices at three sample sizes."
#| fig-width: 8
#| fig-height: 4

set.seed(42)
wishart_demo <- do.call(rbind, lapply(c(20, 50, 200), function(N) {
  samps <- replicate(2000, {
    S <- rWishart(1, df = N - 2, Sigma = diag(3))[,,1]
    diag(S) / (N - 2)
  })
  data.frame(
    s_sq = as.vector(samps),
    N = paste0("N = ", N)
  )
}))

wishart_demo$N <- factor(wishart_demo$N, levels = c("N = 20", "N = 50", "N = 200"))

ggplot(wishart_demo, aes(x = s_sq, fill = N)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  facet_wrap(~ N, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = expression(s[ij]^2 == "diag(S) / (N - 2)"),
       y = "Density") +
  theme(legend.position = "none")
```

As $N$ increases, the distribution concentrates around the true variance (1.0), reflecting reduced sampling variability in the pooled standard deviation.


## Numerical example {#sec-numerical-example}

Let us generate one study's worth of effect sizes to see `generate_smd()` in action. We will create a study with $k = 3$ outcomes, total sample size $N = 100$, true effects $\boldsymbol{\delta} = (0.2, 0.3, 0.15)'$, and within-study correlation $\rho = 0.6$.

```{r}
#| label: example-setup
set.seed(20240601)

# Parameters for one study
delta <- c(0.2, 0.3, 0.15)
k <- 3
N <- 100
rho <- 0.6
```

First, let us construct the equi-correlation matrix and inspect it:

```{r}
#| label: example-sigma
Sigma <- rho + diag(1 - rho, nrow = k)
Sigma
```

Now generate one set of effect sizes:

```{r}
#| label: example-generate
result <- generate_smd(delta = delta, k = k, N = N, Sigma = rho)
result
```

The `g` column contains three Hedges' $g$ estimates, one per outcome. The `var_g` column contains the corresponding sampling variances. Notice that even though the true effects are $\delta = (0.2, 0.3, 0.15)$, the generated $g$ values differ because of sampling variability.

To build intuition about the sampling distribution, let us replicate this 2,000 times:

```{r}
#| label: example-replicate
set.seed(20240601)

reps <- replicate(2000, {
  res <- generate_smd(delta = delta, k = k, N = N, Sigma = rho)
  res$g
}, simplify = FALSE)

g_matrix <- do.call(rbind, reps)
colnames(g_matrix) <- paste0("g", 1:k)
```

The means across replications should be close to the true effects (demonstrating that $g$ is approximately unbiased for $\delta$ when $N$ is moderate):

```{r}
#| label: example-means
colMeans(g_matrix)
```

And the correlation structure of the generated effect sizes reflects the within-study correlation $\rho = 0.6$:

```{r}
#| label: example-cor
cor(g_matrix) |> round(2)
```

The off-diagonal correlations should be close to --- but not exactly --- 0.6, because the sampling correlation among $g$ values depends on the true effects, sample size, and the population correlation in a more complex way.


## Connection to the DSR workflow {#sec-dsr-connection}

In the seven-step Design-Simulate-Report (DSR) workflow that structures this book, `generate_smd()` lives squarely in **Step 3: Generate**. It is the innermost building block of the data-generating process:

1. **`generate_meta()`** (Chapter [-@sec-generate-meta]) loops over studies, creating true effects and drawing study-level random effects.
2. For each study, it calls **`generate_smd()`** to convert true effects into observed Hedges' $g$ values and sampling variances.
3. The resulting dataset of $(g_{ij}, V_{ij})$ pairs is then passed to the estimation functions in Step 4.

The function is deliberately simple --- it handles a single study and returns a tidy tibble. This modular design means we can test and validate the effect-size generation independently from the rest of the simulation machinery. In the next chapter, we will see how `generate_meta()` orchestrates calls to `generate_smd()` across all $J$ studies in a simulated meta-analysis.
