# Running the Simulation {#sec-execution}

```{r}
#| label: setup-execution
#| include: false
source("_common.R")
```

With the factorial design specified and the pipeline functions tested, we are ready for the main event: running thousands of simulation iterations across all parameter conditions. This chapter walks through the driver scripts that orchestrate execution, discusses parallel processing strategies, and --- critically for this course --- explains why you do not need to run the full simulation yourself.

## Overview {#sec-exec-overview}

@pustejovsky2022 run two parallel simulations, one for each empirical specification:

1. **dvcat** --- driven by `RVE-simulation-dvcat.R`, using calibrated parameters from the dependent-variable-category analysis.
2. **Ctype** --- driven by `RVE-simulation-Ctype.R`, using calibrated parameters from the control-group-type analysis.

The two scripts are nearly identical. They share the same design factors, the same filtering logic, the same pipeline function (`simulate_meta()`), and the same three estimation methods. The only difference is which `.Rdata` file is loaded at the top, which determines the empirical objects (`covariates`, `studyID`, `subgroup`, `sample_sizes`, `beta`, and the variance calibration scalars) passed into the pipeline.

Because the scripts are structurally identical, we will annotate `RVE-simulation-dvcat.R` in detail. Everything applies equally to the Ctype counterpart.


## Walkthrough of the driver script {#sec-walkthrough}

The driver script `R/RVE-simulation-dvcat.R` has four logical blocks. Let us trace each one.

### Block 1 --- Load functions and data

```{r}
#| label: block1-load
#| eval: false
source("simulations/simulation-functions.R")
load(file = "simulations/dv-model-data.Rdata")
source_obj <- ls()
```

The `source()` call brings in all helper functions (`generate_smd()`, `generate_meta()`, `estimate_models()`, `simulate_meta()`). The `load()` call restores the calibrated empirical objects into the global environment. The `source_obj <- ls()` line captures the names of all currently loaded objects --- this list is later passed to the parallel cluster so that every worker has access to the same functions and data.

### Block 2 --- Define the parameter grid

```{r}
#| label: block2-design-factors
#| eval: false
set.seed(20210128)

design_factors <- list(
  n_studies = c(30, 60),
  struct = c("CS", "HCS"),
  tau = c(0.05, 0.10, 0.20),
  omega = c(0.00, 0.10),
  rho = c(0.2, 0.4, 0.6, 0.8),
  nu = 9
)

batches <- 34
total_reps <- 4012
```

The seed is set once at the top for reproducibility. The `design_factors` list defines the six factors from @sec-design. The total number of Monte Carlo replications (4,012) is split into 34 batches of `4012 / 34 = 118` iterations each. Batching serves two purposes: it enables parallel distribution across CPU cores, and it provides natural checkpoints for monitoring progress.

### Block 3 --- Filter and assign seeds

```{r}
#| label: block3-filter-seeds
#| eval: false
params <- expand.grid(c(design_factors, list(batch = 1:batches)))
params <- subset(params, struct == "CS" | omega == 0.00)

params$iterations <- total_reps / batches
params$seed <- round(runif(1) * 2^30) + 1:nrow(params)
```

The `expand.grid()` call crosses all factors (including batch) into a full factorial grid. The `subset()` line removes the redundant HCS + non-zero omega conditions, exactly as described in @sec-filtering. Each row then receives a unique seed derived from the master RNG stream, ensuring reproducibility even when rows are dispatched to different workers in unpredictable order.

### Block 4 --- Parallel execution and storage

```{r}
#| label: block4-parallel-execution
#| eval: false
library(Pusto)
cluster <- start_parallel(source_obj = source_obj, setup = "register")

tm <- system.time(
  results <- plyr::mdply(
    params,
    .f = purrr::possibly(simulate_meta, otherwise = data.frame()),
    covariates = covariates, studyID = studyID,
    subgroup = subgroup, sample_sizes = sample_sizes,
    beta = beta,
    tau_covariate = tau_covariate, tau_corr = tau_corr,
    tau_scalar = tau_scalar,
    min_studies = 4, max_effects = 12,
    mods = c("robu", "CHE", "SCE"),
    r_imputed = 0.6,
    .parallel = TRUE
  )
)

save(tm, params, results, session_info, run_date,
     file = "RVE-simulation-dvcat-results.Rdata")
```

Several details are worth highlighting:

- **`plyr::mdply()`** applies a function row-wise over a data frame of parameters, returning a combined data frame. With `.parallel = TRUE`, it dispatches rows to the registered parallel backend.
- **`purrr::possibly()`** wraps `simulate_meta()` so that if any single iteration errors (e.g., a non-convergent model), it returns an empty data frame instead of crashing the entire run.
- **`mods = c("robu", "CHE", "SCE")`** tells the pipeline to fit all three working models --- robust variance estimation via `robumeta`, correlated-and-hierarchical effects via `metafor::rma.mv`, and subgroup correlated effects via `metafor::rma.mv` with a block-diagonal structure.
- **`system.time()`** records wall-clock and CPU time for the full run.
- The final `save()` bundles results, parameters, timing, and session information into a single `.Rdata` file for downstream analysis.


## Parallel processing {#sec-parallel}

Each row of the parameter grid is independent --- a single call to `simulate_meta()` with a unique seed. This makes the simulation embarrassingly parallel: we can distribute rows across CPU cores with no communication between workers.

::: {.panel-tabset}

#### simhelpers / furrr

The modern approach uses the `future` framework with `furrr::future_pmap()`. The `simhelpers` package provides `evaluate_by_row()`, which wraps this pattern with built-in seed management:

```{r}
#| label: parallel-simhelpers
#| eval: false
library(future)
library(furrr)
library(simhelpers)

plan(multisession, workers = parallel::detectCores() - 1)

results <- evaluate_by_row(
  params,
  .f = simulate_meta,
  covariates = covariates,
  studyID = studyID,
  subgroup = subgroup,
  sample_sizes = sample_sizes,
  beta = beta,
  tau_covariate = tau_covariate,
  tau_corr = tau_corr,
  tau_scalar = tau_scalar,
  min_studies = 4,
  max_effects = 12,
  mods = c("robu", "CHE", "SCE"),
  r_imputed = 0.6
)

plan(sequential)  # clean up
```

The `future` backend handles exporting objects to workers automatically, and `evaluate_by_row()` manages per-row seeding via `future.seed`.

#### Original P&T

The original code uses the `Pusto` package (a personal utility library by the first author) to set up a PSOCK cluster and register it with `plyr`:

```{r}
#| label: parallel-original
#| eval: false
library(Pusto)
cluster <- start_parallel(source_obj = source_obj, setup = "register")

results <- plyr::mdply(
  params,
  .f = purrr::possibly(simulate_meta, otherwise = data.frame()),
  covariates = covariates, studyID = studyID,
  subgroup = subgroup, sample_sizes = sample_sizes,
  beta = beta,
  tau_covariate = tau_covariate, tau_corr = tau_corr,
  tau_scalar = tau_scalar,
  min_studies = 4, max_effects = 12,
  mods = c("robu", "CHE", "SCE"),
  r_imputed = 0.6,
  .parallel = TRUE
)

parallel::stopCluster(cluster)
```

The `start_parallel()` function creates a cluster, exports all objects named in `source_obj`, loads required packages on each worker, and registers the cluster with `plyr` via `doParallel`. The `.parallel = TRUE` flag in `mdply()` then dispatches rows across the cluster.

:::


## Seed management {#sec-seeds}

Reproducibility in parallel simulations requires careful seed management. When iterations run on different workers in non-deterministic order, relying on a single global seed is insufficient.

@pustejovsky2022 handle this by assigning each row of `params` its own unique seed:

```{r}
#| label: seed-assignment
#| eval: false
params$seed <- round(runif(1) * 2^30) + 1:nrow(params)
```

The `runif(1)` call draws from the master RNG stream (seeded by `set.seed(20210128)`), and the sequential offset `1:nrow(params)` ensures every row gets a distinct value. Inside `simulate_meta()`, each iteration calls `set.seed()` with its assigned seed before generating data, so results are identical regardless of how rows are distributed across workers or how many cores are available.

This is a simple and effective strategy. The `future` framework offers a more formal approach via L'Ecuyer-CMRG parallel RNG streams (`future.seed = TRUE`), but the per-row seed approach used here achieves the same practical goal. Modern parallelization via `furrr` or `future` uses L'Ecuyer-CMRG streams by default (`furrr_options(seed = TRUE)`), which ensures reproducibility across multiple cores without the manual per-batch seed assignment used here.


## Timing and computational budget {#sec-timing}

The full simulation is computationally intensive. Each call to `simulate_meta()` generates a synthetic dataset, fits three multivariate meta-regression models (including iterative REML estimation), and extracts coefficient tables. Across 4,012 replications and 72 unique conditions (per specification), this amounts to:

$$
72 \times 34 \text{ batches} \times 118 \text{ iterations/batch} \times 3 \text{ models} \approx 867{,}000 \text{ model fits}
$$

per specification, and roughly 1.7 million model fits total across both dvcat and Ctype.

On a modern multi-core workstation, each specification takes several hours. The exact wall-clock time depends on the number of cores, clock speed, and RAM. The `tm` object saved in the results file records the actual elapsed time from the original run.


## Partial execution for the course {#sec-partial}

**You do not need to run the full simulation to learn from this course.** The full 4,012 replications exist to achieve the Monte Carlo precision reported in the paper. For learning purposes, much smaller runs are sufficient:

- **Debugging (10 reps):** Use 10 replications to verify that the pipeline runs without errors and produces output in the expected format. This takes seconds.
- **Learning (100 reps):** Use 100 replications to see meaningful patterns in the performance metrics, even if the estimates are noisy. This takes a few minutes.

To run a reduced version, simply modify the batching parameters:

```{r}
#| label: partial-execution
#| eval: false
batches <- 1
total_reps <- 10    # or 100 for a more informative run

params <- expand.grid(c(design_factors, list(batch = 1:batches)))
params <- subset(params, struct == "CS" | omega == 0.00)
params$iterations <- total_reps / batches
params$seed <- round(runif(1) * 2^30) + 1:nrow(params)
```

For the analysis chapters that follow (@sec-metrics through @sec-reporting), we will use the pre-computed results from the full simulation. This lets everyone work with the same high-precision estimates regardless of local computing resources.


## Loading pre-computed results {#sec-load-results}

The repository includes both sets of pre-computed results in the `data/` directory. Let us load the dvcat results and examine their structure.

::: {.callout-note}
If the data files are missing, run `source("R/download-data.R")` to check which files are needed and where to obtain them.
:::

```{r}
#| label: load-dvcat
load("data/RVE-simulation-dvcat-results.Rdata")
```

The `load()` call restores four objects into the environment: `results`, `params`, `tm`, and `session_info`. The timing object tells us how long the original run took:

```{r}
#| label: timing-dvcat
tm
```

The `params` data frame records the full parameter grid that was executed:

```{r}
#| label: params-dvcat
dim(params)
names(params)
```

The main `results` data frame contains all simulation output. Let us inspect its dimensions and columns:

```{r}
#| label: results-dvcat-dims
dim(results)
names(results)
```

Each row in `results` corresponds to one model coefficient from one replication under one parameter condition. The design-factor columns (inherited from `params` via `mdply()`) are followed by estimation output columns (estimates, standard errors, confidence interval bounds, degrees of freedom, etc.).

We can similarly load the Ctype results:

```{r}
#| label: load-ctype
load("data/RVE-simulation-Ctype-results.Rdata")
dim(results)
```


## Quick sanity checks {#sec-sanity}

Before moving to formal performance evaluation, a few quick checks confirm that the results are well-formed.

How many unique parameter conditions were executed?

```{r}
#| label: sanity-conditions
load("data/RVE-simulation-dvcat-results.Rdata")

results |>
  distinct(n_studies, struct, tau, omega, rho) |>
  nrow()
```

How many estimation methods appear in the output?

```{r}
#| label: sanity-models
results |>
  distinct(mod) |>
  pull(mod)
```

A glimpse at the first few rows:

```{r}
#| label: sanity-head
results |>
  select(1:10) |>
  head(8)
```

If these checks show 72 unique conditions (for dvcat), three estimation methods, and a results table with tens of thousands of rows, the data are ready for the performance analysis in the next chapter.


## Summary

This chapter covered the mechanics of running the @pustejovsky2022 simulation:

- The two driver scripts (`RVE-simulation-dvcat.R` and `RVE-simulation-Ctype.R`) follow identical logic, differing only in the empirical data objects they load.
- Execution relies on row-wise application of `simulate_meta()` across the parameter grid, with `.parallel = TRUE` distributing work across CPU cores.
- Per-row seeds ensure reproducibility regardless of parallelization details.
- The full simulation involves roughly 1.7 million model fits and takes hours on a multi-core machine.
- For this course, we use pre-computed `.Rdata` results so that everyone can proceed to the analysis chapters without waiting for a lengthy computation.

In the next chapter, we turn to the **P** in ADEMP: computing performance metrics from the simulation output and assessing how well each working model performs across the design conditions.


## Exercises {.unnumbered}

::: {.callout-caution title="Exercise 10.1: Design your own extension"}
Suppose you wanted to add a new factor to the simulation: the assumed working correlation used in estimation (currently fixed at $\rho = 0.6$). Add levels $\rho_{\text{assumed}} = 0.4$ and $\rho_{\text{assumed}} = 0.8$ to the design grid. How many total conditions would this create? How many additional replications would be needed?

::: {.callout-tip title="Hint" collapse="true"}
Multiply the current 72 conditions by the 3 levels of $\rho_{\text{assumed}}$. The computational cost scales linearly with the number of conditions.
:::
:::
