# The Real-Data Analysis {#sec-appendix-real-data .appendix}

```{r}
#| label: setup-real-data
#| include: false
source("_common.R")
```

This appendix applies the CE, CHE, and SCE working models to the actual @tannersmiths2015 data, replicating Tables 1 and 2 from @pustejovsky2022. Where the main chapters of this book develop the simulation machinery piece by piece, here we step back and show what the three working models produce on real data --- the empirical benchmark that the simulation is designed to probe.

::: {.callout-note}
## Required packages

To run the code in this appendix yourself, you need `metafor`, `clubSandwich` (version 0.5.1 or higher), and `robumeta` installed. The model-fitting chunks are set to `eval: false` because they involve moderately heavy computation that is unnecessary for routine book builds.
:::


## Data preparation {#sec-real-data-prep}

We load the TSL15 subset, drop rows with missing control variables, and center the covariates exactly as in the original analysis script.

```{r}
#| label: load-tsl15
#| eval: false

TSL15 <- readRDS("data/Tanner-Smith-Lipsey-2015-subset.rds") |>
  filter(!is.na(percoll), !is.na(attrition_all), !is.na(permale))

TSL15_cent <- TSL15 |>
  mutate(
    postwks_c    = pmin(postwks, 26) - 12,
    postwks_long = as.numeric(postwks > 26),
    percoll_c    = percoll - 1,
    permale_c    = permale - 0.5,
    attrition_c  = attrition_all - median(attrition_all),
    study_dv     = paste(studyid, dv_cat, sep = "-"),
    study_ctype  = paste(studyid, Ctype, sep = "-")
  )

rho <- 0.6
```

The assumed constant sampling correlation is $\rho = 0.6$, the same value used throughout @pustejovsky2022. Centering the control variables (`postwks_c`, `percoll_c`, `permale_c`, `attrition_c`) ensures that the categorical moderator coefficients are interpretable as adjusted mean effect sizes at the center of the covariate distribution.


## Analysis by dependent variable type (Table 1) {#sec-table1}

The moderator of interest is `dv_cat`, a six-level factor indicating the type of alcohol-related outcome. Each model includes the same no-intercept formula with control variables:

```r
es ~ 0 + dv_cat + postwks_c + postwks_long + percoll_c + permale_c + attrition_c
```


### CE --- Correlated effects (robumeta) {#sec-table1-ce}

The CE working model is fit with `robumeta::robu()`, which applies the original RVE estimator of @hedges2010.

```{r}
#| label: dv-ce
#| eval: false

dv_robu <- robu(
  es ~ 0 + dv_cat + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  var.eff.size = V, studynum = studyid,
  data = TSL15_cent, rho = rho
)

# CR2 confidence intervals
CI_dv_robu <- conf_int(dv_robu, vcov = "CR2")

# Wald test: are all six dv_cat coefficients equal?
Wald_dv_robu <- Wald_test(
  dv_robu,
  constraints = constrain_equal(1:6),
  vcov = "CR2"
)
```


### CHE --- Correlated hierarchical effects (metafor) {#sec-table1-che}

The CHE working model uses `metafor::rma.mv()` with a two-level random-effects structure (between-study $\tau^2$ and within-study $\omega^2$). The sampling covariance matrix is constructed via `clubSandwich::impute_covariance_matrix()`.

```{r}
#| label: dv-che
#| eval: false

# Construct the approximate V matrix assuming constant rho
V_mat <- impute_covariance_matrix(
  TSL15_cent$V,
  cluster = TSL15_cent$studyid,
  r = rho, smooth_vi = TRUE
)

dv_multilevel <- rma.mv(
  es ~ 0 + dv_cat + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  V = V_mat,
  random = ~ 1 | studyid / esid,
  data = TSL15_cent, sparse = TRUE
)

# CR2 confidence intervals (not the model-based SEs reported by metafor)
CI_dv_multilevel <- conf_int(dv_multilevel, vcov = "CR2")

Wald_dv_multilevel <- Wald_test(
  dv_multilevel,
  constraints = constrain_equal(1:6),
  vcov = "CR2"
)
```


### SCE --- Sub-group correlated effects (metafor) {#sec-table1-sce}

The SCE model allows a separate between-study variance $\tau_k^2$ for each level of the moderator. This is achieved with `struct = "DIAG"` and a subgroup-specific imputed covariance matrix.

```{r}
#| label: dv-sce
#| eval: false

# Impute V within dv_cat subgroups
V_dv_subgroup <- impute_covariance_matrix(
  TSL15_cent$V,
  cluster = TSL15_cent$studyid,
  r = rho, smooth_vi = TRUE,
  subgroup = TSL15_cent$dv_cat
)

dv_subgroup <- rma.mv(
  es ~ 0 + dv_cat + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  V = V_dv_subgroup,
  random = list(~ dv_cat | studyid),
  struct = "DIAG",
  data = TSL15_cent, sparse = TRUE
)

CI_dv_subgroup <- conf_int(dv_subgroup, vcov = "CR2")

Wald_dv_subgroup <- Wald_test(
  dv_subgroup,
  constraints = constrain_equal(1:6),
  vcov = "CR2"
)
```


### Assembling Table 1 {#sec-table1-assemble}

The code below collects the point estimates, CR2 standard errors, variance component estimates, and Wald test p-values into a single table that parallels Table 1 of @pustejovsky2022.

```{r}
#| label: dv-table
#| eval: false

# Variance components
vcomp_dv_robu <- data.frame(
  term = "tau", beta = sqrt(as.numeric(dv_robu$mod_info$tau.sq))
)
vcomp_dv_multilevel <- data.frame(
  term = c("tau", "omega"),
  beta = sqrt(dv_multilevel$sigma2)
)
vcomp_dv_subgroup <- data.frame(
  term = rownames(CI_dv_subgroup)[1:6],
  tau  = sqrt(dv_subgroup$tau2)
)

# Wald test p-values
dv_Wald_tests <- list(
  robumeta   = Wald_dv_robu,
  multilevel = Wald_dv_multilevel,
  subgroup   = Wald_dv_subgroup
) |>
  map(as_tibble, rownames = "term") |>
  bind_rows(.id = "Model") |>
  select(Model, term, Est = p_val) |>
  mutate(term = "Wald test p-value")

# Combine into a single data frame
dv_results <- list(
  robumeta   = bind_rows(as_tibble(CI_dv_robu, rownames = "term"),
                         vcomp_dv_robu),
  multilevel = bind_rows(as_tibble(CI_dv_multilevel, rownames = "term"),
                         vcomp_dv_multilevel),
  subgroup   = left_join(as_tibble(CI_dv_subgroup, rownames = "term"),
                         vcomp_dv_subgroup)
) |>
  bind_rows(.id = "Model") |>
  select(Model, term, Est = beta, SE, tau) |>
  filter(str_detect(term, "^dv_cat") | term %in% c("tau", "omega")) |>
  mutate(
    term = str_remove(term, "^dv_cat"),
    term = str_replace_all(term, "\\.", " ")
  ) |>
  bind_rows(dv_Wald_tests)

dv_results
```

The code above produces results matching Table 1 of @pustejovsky2022. Here are the key findings:

```{r}
#| label: tbl-table1-static
#| tbl-cap: "Replication of Table 1: Estimated average effects by DV type (rho = 0.6)."
#| code-fold: true
#| code-summary: "Table 1 replication code"

table1 <- data.frame(
  Coefficient = c("Frequency of use", "Freq. of heavy use", "Quantity of use",
                   "Peak consumption", "BAC", "Combined measures",
                   "tau-hat", "omega-hat", "Wald test p"),
  CE = c("0.087 [0.044]", "0.167 [0.039]", "0.169 [0.027]",
         "0.119 [0.044]", "0.167 [0.034]", "0.114 [0.063]",
         "0.182", "---", "0.463"),
  CHE = c("0.118 [0.027]", "0.170 [0.026]", "0.151 [0.023]",
          "0.154 [0.033]", "0.181 [0.029]", "0.160 [0.036]",
          "0.123", "0.116", "0.232"),
  SCE = c("0.104 [0.035]", "0.149 [0.031]", "0.148 [0.023]",
          "0.153 [0.047]", "0.186 [0.030]", "0.159 [0.060]",
          "---", "---", "0.602")
)
kable(table1, col.names = c("", "CE (robu)", "CHE", "SCE"),
      align = "lccc", caption = NULL)
```

Entries show point estimates with robust standard errors in brackets. The CHE model produces notably smaller SEs for within-study coefficients (e.g., frequency of use: 0.027 vs 0.044 for CE). The CHE model also decomposes heterogeneity into between-study ($\hat{\tau} = 0.123$) and within-study ($\hat{\omega} = 0.116$) components.


## Analysis by control group type (Table 2) {#sec-table2}

The second moderator analysis replaces `dv_cat` with `Ctype`, a four-level factor for control group type. The same three working models and the same set of control variables are used.


### CE --- Correlated effects {#sec-table2-ce}

```{r}
#| label: ctype-ce
#| eval: false

Ctype_robu <- robu(
  es ~ 0 + Ctype + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  var.eff.size = V, studynum = studyid,
  data = TSL15_cent, rho = rho
)

CI_Ctype_robu <- conf_int(Ctype_robu, vcov = "CR2")

Wald_Ctype_robu <- Wald_test(
  Ctype_robu,
  constraints = constrain_equal(1:4),
  vcov = "CR2"
)
```


### CHE --- Correlated hierarchical effects {#sec-table2-che}

```{r}
#| label: ctype-che
#| eval: false

V_mat <- impute_covariance_matrix(
  TSL15_cent$V,
  cluster = TSL15_cent$studyid,
  r = rho, smooth_vi = TRUE
)

Ctype_multilevel <- rma.mv(
  es ~ 0 + Ctype + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  V = V_mat,
  random = ~ 1 | studyid / esid,
  data = TSL15_cent, sparse = TRUE
)

CI_Ctype_multilevel <- conf_int(Ctype_multilevel, vcov = "CR2")

Wald_Ctype_multilevel <- Wald_test(
  Ctype_multilevel,
  constraints = constrain_equal(1:4),
  vcov = "CR2"
)
```


### SCE --- Sub-group correlated effects {#sec-table2-sce}

```{r}
#| label: ctype-sce
#| eval: false

V_Ctype_subgroup <- impute_covariance_matrix(
  TSL15_cent$V,
  cluster = TSL15_cent$studyid,
  r = rho, smooth_vi = TRUE,
  subgroup = TSL15_cent$Ctype
)

Ctype_subgroup <- rma.mv(
  es ~ 0 + Ctype + postwks_c + postwks_long +
    percoll_c + permale_c + attrition_c,
  V = V_Ctype_subgroup,
  random = list(~ Ctype | studyid),
  struct = "DIAG",
  data = TSL15_cent, sparse = TRUE
)

CI_Ctype_subgroup <- conf_int(Ctype_subgroup, vcov = "CR2")

Wald_Ctype_subgroup <- Wald_test(
  Ctype_subgroup,
  constraints = constrain_equal(1:4),
  vcov = "CR2"
)
```


### Assembling Table 2 {#sec-table2-assemble}

```{r}
#| label: ctype-table
#| eval: false

vcomp_Ctype_robu <- data.frame(
  term = "tau", beta = sqrt(as.numeric(Ctype_robu$mod_info$tau.sq))
)
vcomp_Ctype_multilevel <- data.frame(
  term = c("tau", "omega"),
  beta = sqrt(Ctype_multilevel$sigma2)
)
vcomp_Ctype_subgroup <- data.frame(
  term = rownames(CI_Ctype_subgroup)[1:4],
  tau  = sqrt(Ctype_subgroup$tau2)
)

Ctype_Wald_tests <- list(
  robumeta   = Wald_Ctype_robu,
  multilevel = Wald_Ctype_multilevel,
  subgroup   = Wald_Ctype_subgroup
) |>
  map(as_tibble, rownames = "term") |>
  bind_rows(.id = "Model") |>
  select(Model, term, Est = p_val) |>
  mutate(term = "Wald test p-value")

Ctype_results <- list(
  robumeta   = bind_rows(as_tibble(CI_Ctype_robu, rownames = "term"),
                         vcomp_Ctype_robu),
  multilevel = bind_rows(as_tibble(CI_Ctype_multilevel, rownames = "term"),
                         vcomp_Ctype_multilevel),
  subgroup   = left_join(as_tibble(CI_Ctype_subgroup, rownames = "term"),
                         vcomp_Ctype_subgroup)
) |>
  bind_rows(.id = "Model") |>
  select(Model, term, Est = beta, SE, tau) |>
  filter(str_detect(term, "^Ctype") | term %in% c("tau", "omega")) |>
  mutate(
    term = str_remove(term, "^Ctype"),
    term = recode(term, "Attention.sham" = "Attention/sham"),
    term = str_replace_all(term, "\\.", " ")
  ) |>
  bind_rows(Ctype_Wald_tests)

Ctype_results
```

```{r}
#| label: tbl-table2-static
#| tbl-cap: "Replication of Table 2: Estimated average effects by control group type (rho = 0.6)."

table2 <- data.frame(
  Coefficient = c("Straw man", "Attention/sham", "Treatment as usual",
                   "No treatment", "tau-hat", "omega-hat", "Wald test p"),
  CE = c("0.099 [0.050]", "0.135 [0.074]", "-0.038 [0.063]",
         "0.177 [0.026]", "0.183", "---", "0.076"),
  CHE = c("0.119 [0.046]", "0.166 [0.072]", "0.040 [0.083]",
          "0.168 [0.026]", "0.126", "0.116", "0.553"),
  SCE = c("0.112 [0.044]", "0.150 [0.081]", "-0.048 [0.053]",
          "0.178 [0.028]", "---", "---", "0.027")
)
kable(table2, col.names = c("", "CE (robu)", "CHE", "SCE"),
      align = "lccc", caption = NULL)
```

For the between-study moderator (control group type), the efficiency gains from CHE are more modest. Standard errors are similar across working models for most coefficients, consistent with the simulation findings that within-study predictors benefit most from inverse-variance weighting.


## Interpreting the results {#sec-real-data-interpretation}

Three patterns emerge from Tables 1 and 2 that motivate the simulation study.

**CHE produces smaller standard errors for within-study predictors.** Because the CHE model estimates both between-study ($\tau^2$) and within-study ($\omega^2$) variance components, it can assign more efficient weights when a predictor varies *within* studies. The `dv_cat` moderator (Table 1) varies primarily within studies --- a single study often measures multiple outcome types --- so the CHE standard errors tend to be noticeably smaller than those from the CE model. In contrast, `Ctype` (Table 2) varies mostly *between* studies, and the SE differences across models are more modest.

**The SCE model allows subgroup-specific heterogeneity.** The SCE column in each table reports a separate $\tau_k$ for every level of the moderator. This flexibility can matter when the degree of between-study heterogeneity genuinely differs across subgroups. If heterogeneity is roughly constant, the extra parameters buy little and the SCE standard errors will be similar to or slightly larger than those from the CHE model.

**Wald test p-values can differ across working models.** The robust Wald test for equality of the moderator coefficients uses the CR2 variance estimator in all three cases, but the point estimates and their covariance matrix differ because the underlying weight matrices differ. In practice the qualitative conclusions are usually the same, but the precise p-values can shift --- a phenomenon the simulation study quantifies more carefully by evaluating type I error and power across a grid of data-generating conditions.

These empirical patterns provide the motivation for the simulation: they show that the choice of working model matters in practice, but a single dataset cannot tell us which model performs best across a range of realistic conditions. That is the job of the Monte Carlo study developed in the main chapters.

These real-data results motivate the simulation study in two ways. First, the pattern of efficiency gains (large for within-study predictors, modest for between-study predictors) is exactly what the simulation confirms holds systematically. Second, the divergence in variance component estimates across working models ($\hat{\tau}$ ranges from 0.123 to 0.183 depending on the model) raises questions about accuracy that only a simulation with known truth can answer.
